A neutrality index for quantitative traits

### 2023-01-13 manuscript/main.tex
Draft of the manuscript for JEB.

\documentclass{article}
\usepackage{pdfpages}    % To import the PDF document
\usepackage{xcolor}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTBLUELINK}{HTML}{3366BB}
\definecolor{PURPLELINK}{HTML}{663366}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false ]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
linkcolor=DARKBLUELINK,
anchorcolor=DARKBLUELINK,
citecolor=DARKBLUELINK,
filecolor=DARKBLUELINK,
menucolor=DARKBLUELINK,
urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\definecolor{RED}{HTML}{EB6231}
\definecolor{YELLOW}{HTML}{E29D26}
\definecolor{BLUE}{HTML}{5D80B4}
\definecolor{LIGHTGREEN}{HTML}{6ABD9B}
\definecolor{GREEN}{HTML}{8FB03E}
\definecolor{PURPLE}{HTML}{BE1E2D}
\definecolor{BROWN}{HTML}{A97C50}
\definecolor{PINK}{HTML}{DA1C5C}

\usepackage[margin=60pt]{geometry}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{blkarray}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{verbatim}
\usepackage{float}
\usepackage{bm}
\usepackage{listings, enumerate, enumitem}
\lstset{%
	backgroundcolor=\color{gray!25},
	basicstyle=\ttfamily,
	breaklines=true,
	columns=fullflexible
}
\usepackage[export]{adjustbox}
\usepackage{tabu}
\tabulinesep=0.6mm
\newcommand\cellwidth{\TX@col@width}
\usepackage{hhline}
\setlength{\arrayrulewidth}{1.2pt}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\AtBeginEnvironment{tabu}{\footnotesize}
\usepackage{booktabs}

\usepackage{graphicx}
\graphicspath{{artworks/}}
\makeatletter
\def\input@path{{artworks/}}
\makeatother
\pdfstringdefDisableCommands{%
\renewcommand*{\bm}[1]{#1}%
% any other necessary redefinitions
}
\usepackage{xfrac, nicefrac}
\usepackage{lmodern}
\usepackage{natbib}
\pdfinclusioncopyfonts=1

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\angstrom}{\text{\normalfont\AA}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Pfix}{2 \Ne \proba_{\text{fix}}}

\newcommand{\VarPhenotype}{V_{P}}
\newcommand{\VarGenetic}{V_{G}}
\newcommand{\MatrixGenetic}{\BiDimArray{G}}
\newcommand{\VarMutation}{V_{M}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}
\newcommand{\MutationRate}{\mu}
\newcommand{\NbrLoci}{L}
\newcommand{\Trait}{X}

@@

\part*{Evolutionary quantitative genomics}
\tableofcontents

\section{Quantitative genetics theory}

In this section, we seek to derive phenotypic and genetic variance as a function of mutation, selection and drift.
First, the genetic variance ($\VarGenetic$) is related to the observed phenotypic variance ($\VarPhenotype$) through heredity:
\begin{gather}
    \VarGenetic =  h^2 \VarPhenotype \label{eq-heredity}.
\end{gather}

\subsection{Mutation-drift equilibrium}

Mutation supply new variants in the population, while genetic random drift depletes standing variation in the population. At equilibrium between mutation and drift~\citep{lynch_mutation_1998}, the genetic variance is thus a function of the mutational variance ($\VarMutation$) and the effective number of individual in the population ($\Ne$):
\begin{gather}
    \VarGenetic =  2 \Ne \VarMutation \label{eq-var-genetic}.
\end{gather}

The mutational variance ($\VarMutation$) is the rate at which new mutations contributes to new mutations. As shown in \citet{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ depends on the mutation rate per loci per generation ($\MutationRate$), the number of loci encoding the trait ($\NbrLoci$) and finally the random effect on the phenotype of a mutation ($a$):
\begin{gather}
    \VarMutation =  2 \NbrLoci \MutationRate \E \left[ a^2 \right] \label{eq-var-mutation}.
\end{gather}

\subsection{Mutation-selection-drift equilibrium}

Selection, whether positive or negative, depletes standing variation in the population and reduces to genetic variance. Hence, at equilibrium between mutation, selection and drift, the observed genetic variance is lower than the theoretical neutral variance:
\begin{align}
    \VarGenetic & < 2 \Ne \VarMutation \\
    \iff \VarGenetic & < 4 \Ne \NbrLoci \MutationRate \E \left[ a^2 \right]
\end{align}

Under the House of cards model of stabilizing selection~\citep{kingman_simple_1978}, the trait's fitness is normally distributed around an optimal value ($\theta$) and with standard deviation $\StdSelection_S$. Moreover, the trait is encoded by a few loci with large effects and the genetic variance at equilibrium between mutation, selection and drift is: 
\begin{gather}
    \VarGenetic = 2 \NbrLoci \MutationRate \VarSelection_S \label{eq-mutsel-HC}.
\end{gather}

Alternatively, under the Gaussian model of stabilizing selection~\citep{lande_natural_1976}, the trait's fitness is also normally distributed around an optimal value ($\theta$) and with standard deviation $\StdSelection_S$. However, the trait is encoded by a large number of loci traits with small effect and the genetic variance at equilibrium is: 
\begin{gather}
    \VarGenetic = \sqrt{2 \NbrLoci \VarMutation \VarSelection_S} \label{eq-mutsel-G}.
\end{gather}

The boundary between the two models is given by \citet{turelli_heritable_1984} as:
\begin{align}
    20 \MutationRate \VarSelection_S &< \E \left[ a^2 \right] \text{ for the House of card model}, \\
    20 \MutationRate \VarSelection_S &> \E \left[ a^2 \right] \text{ for the geometric model}.
\end{align}

\section{Quantitative traits}

By definition, quantitative trait is an observable can be measured for an organisms, for example genome size~\citep{alfsnes_genome_2017}.

\subsection{Life history traits}

Life history traits are a subset of quantitative traits related to the age and stage-specific patterns and timing of events that make up an organism's life such as such as age at maturity, fecundity, life expectancy, height, weight, and so on. Life history traits are often suspected to be under selection and encoded by many loci.

\subsection{Genetic architecture of quantitative traits}
Due to recent advances in Genome Wide Association Studies (GWAS), the availability of  datasets in human (\textit{e.g.} UK Biobank) allows to dissect the genetic underpinning of quantitative traits such as Body Mass Index (BMI) or disease related susceptibility.
For example by relating observable summary statistics (phenotypic distribution, allelic frequencies, variance per site) to parameters of the genetic architecture (pleiotropy, distribution of selection coefficients, effect size), as in \citet{simons_population_2018}.
Overall, as reviewed from GWAS evidences in \citet{sella_thinking_2019}, most quantitative traits are polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and under stabilizing selection in humans.

\subsection{Gene expression}

In transcriptomic dataset, the expression level of each gene can be seen a quantitative trait.
First, $\VarGenetic$ can be estimated from the standing variation inside natural populations. 
Second, $\VarMutation$ can be  estimated from mutation accumulation experiments in laboratory clonal lines.
Third, $\NbrLoci$ can be  estimated from effect of gene knockout of expression level.
Forth, $\MutationRate$ can be  estimated from the frequency of mutation altering expression level.
Fifth, $\Ne$ can be estimated from standing genetic polymorphism ($\pi = 4 \Ne u$), by correcting for the mutation rate ($u$).
Altogether, the evolutionary regime (neutral or selection) of gene expression can be tested across all genes, as suggested in \citet{fay_evaluating_2008}.

In the yeast \textit{Saccharomyces cerevisiae}, the fruit fly \textit{Drosophila melanogaster} and the nematode \textit{Caenorhabditis elegans},  \citet{hodgins-davis_gene_2015} found that gene expression is under stabilizing selection, such that neutral evolution is ruled out for most genes.
Moreover, the House of card model is preferred against the Gaussian model of stabilizing selection, also suggesting that gene expression is encoded by a few loci with large effects.

Finally, mutation accumulation experiments with different population size have shown that $\VarMutation$ decreases with $\Ne$, suggesting that gene expression is under strong stabilizing selection~\citep{deiss_global_2021}.

\section{Phylogenetic trait evolution}

\subsection{Relationship between micro and macro evolution. }

We denote $\Trait_i$ the quantitative trait $\Trait$ for species $i$.
\citet{hansen_translating_1996} derived a relation between trait variation between species and the genetic variance of this trait as 
\begin{gather}
    \Cov \left( \Trait_i, \Trait_j \right) = \Cov \left( \E \left[ \Trait_i | \Trait_z \right ], \E \left[ \Trait_j | \Trait_z \right ] \right),  \label{eq-cov-traits}
\end{gather}
where $z$ is the ancestor between species $i$ an $j$.

If the trait is neutral, evolving as a Brownian process, then $\E \left[ \Trait_i | \Trait_z \right ] = \Trait_z$ and equation \ref{eq-cov-traits} simplifies to:
\begin{align}
\Cov \left( \Trait_i, \Trait_j \right) &= \Var \left( \Trait_z \right),\\
    &= \frac{\VarGenetic}{\Ne} t_{z}, \\
    &=  2 t_{z} \VarMutation, \label{eq-cov-traits-neutral} \text{ from equation } \ref{eq-var-genetic},
\end{align}
where $t_{z}$ is the time between species $z$ an both $i$ an $j$.

The relationship between trait variation between species and the mutational variance can be used as method to detect whether traits are evolving under a neutral model of evolution, similar to the use of synonymous ($\ds$) and non-synonymous ($\dn$) substitution rates in molecular evolution, as noted in \citet{fay_evaluating_2008}.

\begin{table}[H]
	\centering
	\noindent\adjustbox{max width=\textwidth}{%
		\begin{tabu}{|c||c|c|}
			\hline
			 & Coding sequences & Quantitative traits \\ \hline \hline
			Adaptation or moving optimum & $\dnds > 1$ & $\Cov \left( \Trait_i, \Trait_j \right) > 2 t_{z} \VarMutation $ \\ \hline
			Neutral regime of evolution & $\dnds = 1$ & $\Cov \left( \Trait_i, \Trait_j \right) = 2 t_{z} \VarMutation $ \\ \hline
			Purifying or directional selection & $\dnds < 1$ & $\Cov \left( \Trait_i, \Trait_j \right) < 2 t_{z} \VarMutation $ \\ \hline
		\end{tabu}}
\end{table}

If the trait is under directional selection, but the strength and direction of selection varies randomly from one generation to the next (variance $\VarSelection_D$), then the trait still follows a Brownian process model and equation \ref{eq-cov-traits} reduces to:
\begin{align}
\Cov \left( \Trait_i, \Trait_j \right) &= \Var \left( \Trait_z \right), \\
    &= \left(\frac{\VarGenetic}{\Ne} + \VarGenetic \VarSelection_D \VarGenetic \right)  t_{z}. \label{eq-selection-vary-direct}
\end{align}

If the trait is under stabilizing selection, but the strength and direction of the optimum varies randomly from one generation to the next (variance $\VarSelection_{\theta}$), then the trait still follows a Brownian process model and equation \ref{eq-cov-traits} reduces to:
\begin{align}
\Cov \left( \Trait_i, \Trait_j \right) &= \VarSelection_{\theta} t_{z}. \label{eq-selection-vary-stabi}
\end{align}

Altogether, traits that are found to be under a long term Brownian process are not necessarily neutral.

To read : \citet{arnold_adaptive_2001}
\subsection{Trait evolution under a Brownian or Ornstein-Uhlenbeck process}

Modelling $\Trait$ as a Brownian process allows theoretically to test for neutrality.
In \citet{catalan_drift_2019}, gene expression is modelled either as a Brownian (supposedly neutral regime), or as a Brownian with a shift in the mean for one branch (supposedly directional selection), or as an Ornstein-Uhlenbeck process (supposedly stabilizing selection) or finally as an Ornstein-Uhlenbeck process with a shift in the optimal for one branch (supposedly directional selection). 
The analysis on the transcriptome across different individuals and species concluded that 81\% of genes are under a neutral regime and 16\% under directional selection. 
However, the alternative model of a varying directional (equation~ \ref{eq-selection-vary-direct}) or stabilizing selection (equation~ \ref{eq-selection-vary-stabi}) was not implemented, and thus can not be ruled out by this analysis.

\subsection{Variance evolution}

The multidimensional generalization of $\VarGenetic$ to several traits is the of the genetic covariance matrix $\MatrixGenetic$.
Stability of $\MatrixGenetic$, can be reformulated as whether $\MatrixGenetic (t) = \MatrixGenetic$, an assumption discussed in \citet{arnold_understanding_2008} and \citet{hohenlohe_mipod_2008}. Having a time dependent matrix is implemented in JIVE, a Bayesian hierarchical model developed by \citet{kostikova_bridging_2016} and implemented in BEAST by \citet{gaboriau_multi-platform_2020}.

\section{Discussion \& perspectives}
At the population level, gene expression is shown to be empirically under stabilizing selection. However at the phylogenetic scale, gene expression is under a neutral regime. Can we reconcile these observations?

We can find the genes under adaptive regime (fluctuating optimal gene expression) at the phylogenetic scale and compare that to the genes under adaptation using $\dnds$ based methods. Do we find the same genes?
\section{Appendix}

\subsection{Definitions}

We denote $\Trait_{i}$ the quantitative trait $\Trait$ for individual $i$.

\begin{align}
    \VarPhenotype = \sum_{i=1}^{\Ne} \Trait_{i}^2 - \left( \sum_{i=1}^{\Ne} \Trait_{i} \right)^2
\end{align}

\bibliographystyle{natbib}%%%%Bibliography style file
\bibliography{references}

\end{document}

>>>>

%! BibTeX Compiler = biber
%TC:ignore
\documentclass{article}

\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
linkcolor=DARKBLUELINK,
anchorcolor=DARKBLUELINK,
citecolor=DARKBLUELINK,
filecolor=DARKBLUELINK,
menucolor=DARKBLUELINK,
urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage{biorxiv}
\usepackage[backend=biber,eprint=false,isbn=false,url=false,intitle=true,style=nature,date=year]{biblatex}
\addbibresource{references.bib}

\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{xfrac, nicefrac}
\usepackage{bm}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{graphicx}
\usepackage{bbold}
\usepackage{pdfpages}
\pdfinclusioncopyfonts=1
\usepackage{lineno}
\renewcommand{\baselinestretch}{1.25}
\renewcommand{\arraystretch}{1.6}

\input{notations.tex}

\title{A neutrality index for quantitative traits}

\author{
\large
\textbf{T. {Latrille}$^{1}$, T. {Gaboriau}$^{1}$, N. {Salamin}$^{1}$}\\
\normalsize
$^{1}$Université de Lausanne, Lausanne, Switzerland\\
\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}

@@

\maketitle

% Abstract (≤ 250 words)
%TC:endignore
\begin{abstract}
    Current methods at the population scale do not integrate the changes over long evolutionary time, while at the same time the methods require laboratory clonal lines to determine the neutral expectation.
    Conversely, at the phylogenetic scale, current methods discard the variance in traits between individuals, leaving out valuable information.
    The main goal of this project is to combine within-species and between-species variation of quantitative traits.
    To this aim, we define a neutrality index for of a quantitative trait, devoted to determine its evolutionary regime, leveraging both data at the population and phylogenetic scale.
\end{abstract}

\keywords{Quantitative genetics \and Phylogenetics \and Population-genetics \and Selection }

% Research Article (7500 words)
\section{Introduction}\label{sec:introduction}
A quantitative trait is by definition an observable can be measured for an organisms, for example genome size\cite{alfsnes_genome_2017}.
Life history traits (LHTs) are such quantitative traits, related to the age and stage-specific patterns and timing of events that make up an organism's life such as age at maturity, fecundity, life expectancy, height, weight, and so on.
LHTs are often suspected to be under selection and encoded by many loci.
Appart from LHTs, the genetic underpinning of other quantitative traits such as Body Mass Index (BMI) or disease related susceptibility can now be dissected due to recent advances in Genome Wide Association Studies (GWAS) and the availability of datasets in human (\textit{e.g.} UK Biobank)\cite{simons_population_2018}.
Overall, GWAS have revealed that traits are polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and under stabilizing selection in humans\cite{sella_thinking_2019}.
At the genomic level, the expression level of each gene can also be seen as a quantitative trait.
Surveys of gene expression reveal extensive variability both within and between a wide range of species\cite{fay_evaluating_2008}.

From the standing variation both within and between a wide range of species, distinguishing adaptive changes driven by positive selection from neutral evolution is critical for understanding the evolution of trait evolution.
At the population level, the variance between individuals of the same population can be estimated for any quantitative trait.
This observed variance can be compared to the expected variance that would be obtained under the assumption that the trait is not under selection, which can be estimated from mutation accumulation experiments in laboratory clonal lines.
At the phylogenetic scale, between different populations, the evolutionary regime for a quantitative trait can also be tested.
Typically, the trait mean (across individuals of the same population) is modelled as a Brownian process, allowing theoretically to test for neutrality.
Alternatively, a Brownian with a shift in the mean for one branch is supposedly directional selection, an Ornstein-Uhlenbeck process is supposedly stabilizing selection or finally an Ornstein-Uhlenbeck process with a shift in the optimal for one branch is supposedly directional selection\cite{catalan_drift_2019}.

At the frontier between phylogenetics and population genetics, studies comparing between species variation have acknowledged the importance of within species variation\cite{whitehead_variation_2006}.
However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within populations is a promising line of research\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
Altogether, there is a push toward integrating within and between species variation into a unified framework.
This relationship had been proposed to describe the evolutionary regime of a trait for a pair of species\cite{lande_genetic_1980, turelli_heritable_1984}.
However, it has not yet been generalized for any number of species related by their phylogenetic tree.
Finally, studies on quantitative trait evolution
In this study, we propose a neutrality index for a quantitative trait, leveraging data within and between species, normalizing the trait evolution by its genetic polymorphism and divergence.

\section{Materials and Methods}\label{sec:materials-and-methods}
\subsection{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
\subsubsection{Mutation-drift equilibrium for a trait}

For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect on the random of a mutation ($a$).
New mutations are thus generating variance for such a trait.
We define the expected effect of a mutation on the variance of a trait as $\RateMut$, given as:
\begin{gather}
    \RateMut \defEqual \NbrLoci \Multiply \E \left[ a^2 \right]. \label{eq:rate-mut}
\end{gather}

At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contributes to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of $\RateMut$ and the mutation rate per loci per generation ($\MutationRate$), as:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

At the population level, mutations supply new variants in the population, while genetic random drift depletes standing variation.
As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in the population ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individual in the population ($\Ne$) as:
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

Moreover, for any neutral genomic region of interest, the genetic diversity, called $\pi$, is measured as the number of mutation segregating in the population divided by the size of the region.
Any of the segregating mutation will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is thus a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$), given as:
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

Altogether, at the population level, we define define $\RatePop$ as the ratio of genetic variance of the trait ($\VarGenetic$) over genetic diversity of any neutral genomic region of interest ($\pi$), which identify with $\RateMut$:
\begin{align}
    \RatePop & \defEqual \frac{\VarGenetic }{\pi}, \\
    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and \ref{eq:genetic-diversity}}, \\
    & = \RateMut. \label{eq:rate-pop}
\end{align}

Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heredity ($h^2$), giving also $\RatePop$ as:
\begin{align}
    \RatePop & = \frac{h^2 \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

\subsubsection{Phylogenetic evolution of a trait}

Along a lineage, we denote $\Trait$ as the mean value of the trait over the individuals in the population.
If the trait is neutral and encoded by many loci, $\Trait$ evolves as Brownian process\cite{hansen_translating_1996}.
After the population has evolved for $\Time$ generations, the variance of $\Trait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated in the populations, some will we eventually reach fixation in the population due to genetic drift, resulting in a substitution.
For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate with which a new neutral allele is fixed in the population in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region\cite{kimura_evolutionary_1968}.

Thus, after having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRate \text{ from eq.\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
Altogether, the variance in the mean trait value ($\VarPhy$) normalized by the genetic distance of any neutral genomic region of interest ($d$) also identify with $\RateMut$:
\begin{align}
    \RatePhy & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and \ref{eq:distance}}, \\
    & = \RateMut \label{eq:rate-phy}.
\end{align}

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-summary}
    \caption{
        In red, at the phylogenetic scale, the variance of mean trait value ($\VarPhy$) normalised by genetic distance ($d$) is defined as $\RatePhy$.
        In blue, at the population scale, trait variance within species ($\VarGenetic$) normalised by genetic diversity ($\pi$) is defined as $\RatePop$.
        Under neutral evolution $\RatePhy$ should equal $\RatePop$.
        Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
    }
    \label{fig:methods}
\end{figure*}

\subsubsection{Neutrality index}

 The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
 At the population level, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
Combined as a ratio, we obtain $\RatePop$ at the population level, which is a proxy for $\RateMut$, namely the expected effect of a mutation on the variance of a trait.
At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allow to compute $d$.
Similarly, combined as a ratio, we obtain $\RatePhy$ which is another proxy for $\RateMut$.
Altogether, we thus have:
\begin{align}
    \RateMut & = \RatePop = \RatePhy \text{ from eq.~\ref{eq:rate-pop} and \ref{eq:rate-phy}} , \\
    \Rightarrow \frac{\RatePhy}{\RatePop} & = \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d }  = 1.
\end{align}
Altogether, $\RatePhy / \RatePop$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
Importantly, $\RatePhy$ and $\RatePop$ can both be estimated using quantitative trait and genomic sequences within and between populations, while neither the mutation rate $\MutationRate$ nor the effective population size $\Ne$ need to be estimated.

\subsection{Estimates}

\subsubsection{Maximum likelihood estimate}
For each population of interest with inter-individual data available, $\RatePop$ can be estimated as the ratio of genetic variance of a trait over genetic diversity of neutrally evolving sequences thus normalizing .
Averaged across all populations we obtain the mean as:
\begin{gather}
    \EstRatePop = \left< \frac{\VarGenetic}{\pi} \right> = \left< \frac{h^2 \Multiply \VarPhenotype }{\pi } \right> \label{eq:estimated-rate-pop}
\end{gather}

At the phylogenetic scale, for $\NbrSpecies$ taxa in the tree, $\Cr$ is the $\NbrSpecies \times \NbrSpecies$ contrast matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
The contrast matrix is symmetric, along the diagonal are the total distances ($d_{i}$) from the root of the tree to each taxon ($i$).
The off-diagonal elements $C_{\Spi,\Spj}$ are the total branch lengths shared by particular pairs of taxa $\Spi$ and $\Spj$, computed from the distance between the pair ($d_{\Spi,\Spj}$) as:
\begin{gather}
    C_{\Spi,\Spj} = d_{\mathcal{A}(\Spi, \Spj)} = \frac{d_{\Spi} + d_{\Spj} - d_{\Spi,\Spj}}{2}
\end{gather}
Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrSpecies \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate\cite{omeara_testing_2006}, given as:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \Cr\inv \VecOne \right)\inv \left( \VecOne\tr \Cr\inv \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrSpecies \times 1$ column vector of ones.

Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated\cite{omeara_testing_2006} as:
\begin{gather}
    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \VecOne \right)\tr \Cr\inv \left( \VecTrait -  \RootTrait  \VecOne  \right)}{\NbrSpecies - 1}. \label{eq:estimated-rate-phy}
\end{gather}
Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.

\subsection{Simulations}\label{subsec:simulations}
The evolutionary dynamics was formalized as a Wright-Fisher model with mutation, selection and drift.
The population is assumed to be panmictic, with {effective population size} $\Ne$ and with non-overlapping generations.
We simulate explicitly each generation along the phylogeny under a Wright-Fisher population, consisting of three steps: mutation, selection and {genetic drift} of currently segregating alleles.
Mutations are drawn based on a user-defined nucleotide matrix, where our simulations used a symmetric time-reversible mutation matrix.
Drift is induced by the multinomial resampling of the currently segregating alleles.
Our simulator use a geometric Brownian multivariate process to model the changes in the mutation rate per generation, the generation time and $\Ne$ along the lineages.
Moreover, the instant value of log-$\Ne$ can also be modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
In our simulations, the tree is composed of 77 species (see supplementary), the tree root is $150$ million years old, the initial mutation rate is $\smash{10^{-8}}$ per site per generation and the initial generation time is $10$ years.
The simulator starts from an initial sequence at equilibrium, at each node, the process is split, and finally stopped at the leaves of the tree.

We introduced selection as a one dimensional Fisher's geometric landscape~\cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
The phenotype is a vector in a multidimensional space, where the number of dimensions is often termed complexity.
From a {phenotype}, the fitness is a monotonously decreasing function of the {phenotype} distance to $0$.
The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength ($\alpha$) and epistasis ($\beta$).
If the phenotype-fitness map is explicit, the genotype-phenotype map is more pervasive.
Mutations are seen has displacement of the {phenotype} in the multidimensional space.
Beneficial mutations are moving the {phenotype} closer to $0$, whereas deleterious mutations are moving the {phenotype} further away.
In such original context, the distribution of mutational effects is not dependent on the current genotype, but this can be relaxed using a genotype-phenotype map.

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-simulator}
    \caption{
        Wright-Fisher simulations with mutation, selection and drift.
    }
    \label{fig:simulator}
\end{figure*}

\section{Results}

To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at the population and phylogenetic scale.
Quantitative genetics formalism has related both within and between species variation to the underlying mutational input for a neutral trait.
More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$) is a proxy for the mutational variance ($\RatePop$).
At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a time $\Time$ is called $\VarPhy$.
This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), allowing us to obtain another proxy for the mutational variance ($\RatePhy$).
Altogether, we define the neutrality index as $\RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
Under a depleted variation between species, meaning $\RatePhy < \RatePop$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
In contrast, under an inflated variation between species, meaning $\RatePhy > \RatePop$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.
